---
title: "earlyR Testing"
author: "Christine Sangphet"
date: "2024-06-25"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Worked Example

### Assume a small outbreak of Ebola Virus Disease (EVD), for which the serial interval has been previously characterised. We study a fake outbreak, for which we will quantify infectiousness (R), and then project future incidence

```{r}
#load environment 

library(earlyR)
library(ggplot2)
library(incidence)
```

```{r}
#define the onset dates

onset <- as.Date(c("2017-02-04", "2017-02-12", "2017-02-15",
                   "2017-02-23", "2017-03-01", "2017-03-01",
           "2017-03-02", "2017-03-03", "2017-03-03")) 
```

```{r}
#compute the daily incidence

library(incidence)

i <- incidence(onset)
i

plot(i, border = "white")
```

```{r}
#assume the current date is 3/21/2017. we need to indicate this to incidence using:

today <- as.Date("2017-03-21")
i <- incidence(onset, last_date = today)
i

#plot and adjust x-axis

plot(i, border = "white") +
  scale_x_date(
    breaks = seq(min(i$dates), max(i$dates), by = "week"),
    labels = function(x) format(x, "%Y/%m/%d")
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#note: it is essential to make sure that the last days without cases are included (omitting this information would lead to an over-estimation of R)
```

### Estimating the Likelihood Distribution of R

```{r}
#for estimating R, we need estimates of the mean and standard deviation (sd) of the serial interval i.e. the delay between the primary and secondary sympton onset dates

#define mean and sd (defined on website)

mu <- 15.3 # mean in days days
sigma <- 9.3 # sd in days 

#the function get_R is then used to estimate the most likely values of R:

res <- get_R(i, si_mean = mu, si_sd = sigma)
res

plot(res)
```

### Represent infectiousness over time

```{r}
plot(res, "lambdas", scale = length(onset) + 1) +
  geom_vline(xintercept = onset, col = "grey", lwd = 1.5) +
  geom_vline(xintercept = today, col = "blue", lty = 2, lwd = 1.5)

#the figure shows the global force of infection over time, with vertical grey bards indicating the dates of sympton of onset. the dashed blue line indicates current day. note: the vertical scale for the bars is arbitrary, and only represents the relative force of infection
```

### Using sample_R to get a large sample of likely R values

```{r}
R_val <- sample_R(res, 1000)
summary(R_val)

quantile(R_val)

quantile(R_val, c(0.025, 0.975))

hist(R_val, border = "grey", col = "navy",
     xlab = "Values of R",
     main = "Sample of likely R values")
```

### Simulating future epicurves

```{r}
#retrieve the serial interval (SI)

si <- res$si
si
```

```{r}
#use the package projections 

library(projections)

future_i <- project(i, R = R_val, n_sim = 1000, si = res$si, n_days = 30)
future_i

mean(future_i) 
plot(future_i)
```

### Having the distribution of the total number of new cases predicted in the next 30 days

```{r}
predicted_n <- colSums(future_i)
summary(predicted_n)
hist(predicted_n, col = "darkred", border = "white",
     main = "Prediction: new cases in 30 days",
     xlab = "Total number of new cases")
```

### What the results will look like without the recent zero incidence

```{r}
alt_i <- incidence(onset)
alt_res <- get_R(alt_i, si_mean = mu, si_sd = sigma)
alt_R_val <- sample_R(alt_res, 1000)
alt_future_i <- project(alt_i, R = alt_R_val, n_sim = 1000, si = res$si, n_days = 30)
alt_future_i

mean(alt_future_i)
#> [1] 5.5434
plot(alt_future_i)
```

```{r}
## alternative plot
col <- "#cc66991a"
matplot(alt_future_i, type = "l", col = col, lty = 1, lwd = 5,
        xlab = "Day from today",
    ylab = "Projected daily incidence")
```

```{r}
alt_predicted_n <- colSums(alt_future_i)
summary(alt_predicted_n)

hist(alt_predicted_n, col = "darkred", border = "white",
     main = "Prediction: new cases in 30 days",
     xlab = "Total number of new cases")

```
