---
title: "EpiEstim x earlyR testing"
author: "Christine Sangphet"
date: "2024-07-19"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **EpiEstim x earlyR**

```{r}
#load environment

library(EpiEstim)
library(ggplot2)
library(incidence)
library(earlyR)
library(dplyr)
```

```{r}
#load data

data(Flu2009)
```

### **Estimations with EpiEstim**

```{r}
res_parametric_si <- estimate_R(Flu2009$incidence, 
                                method="parametric_si",
                                config = make_config(list(
                                  mean_si = 2.6, 
                                  std_si = 1.5))
)
```

```{r}
plot(res_parametric_si)
```

### **Estimations with earlyR**

```{r}
Flu2009$incidence
```

```{r}
dates <- c("2009-04-27", "2009-04-28", "2009-04-29", "2009-04-30",
           "2009-05-01", "2009-05-02", "2009-05-03", "2009-05-04",
           "2009-05-05", "2009-05-06", "2009-05-07", "2009-05-08",
           "2009-05-09", "2009-05-10", "2009-05-11", "2009-05-12",
           "2009-05-13", "2009-05-14", "2009-05-15", "2009-05-16",
           "2009-05-17", "2009-05-18", "2009-05-19", "2009-05-20",
           "2009-05-21", "2009-05-22", "2009-05-23", "2009-05-24",
           "2009-05-25", "2009-05-26", "2009-05-27", "2009-05-28")

counts <- c(1, 1, 0, 2, 5, 3, 3, 3, 6, 2, 5, 9, 13, 12, 13, 11, 12, 6, 6, 6, 
            3, 1, 0, 2, 0, 0, 0, 0, 2, 0, 2, 0)

onset <- as.Date(rep(dates, counts))
```

```{r}
i <- incidence(onset)
i
```

```{r}
plot(i, border = "white")
```

#### R

```{r}
mu <- 2.6 # mean in days days
sigma <- 1.5 # sd in days 

#the function get_R is then used to estimate the most likely values of R:

res <- get_R(i, si_mean = mu, si_sd = sigma)
res
```

```{r}
plot(res)
```

#### SI

```{r}
# Parameters from the serial interval distribution

shape <- 3.00444444444445
scale <- 0.865384615384615

# Generate a sequence of values

x <- seq(0, 10)

# Calculate the density of the gamma distribution

y <- dgamma(x, shape = shape, scale = scale)

# Create a data frame for plotting

si_df <- data.frame(x = x, y = y)

# Plot using ggplot2

si_earlyR <- ggplot(si_df, aes(x = x, y = y)) +
  geom_line() +
  labs(title = "Serial Interval Distribution",
       x = "Time",
       y = "Density") +
  theme_minimal()

print(si_earlyR)
```

### Comparison of SI estimation

#### Side-by-side

```{r}
# Plot for res_parametric_si
plot1 <- plot(res_parametric_si, "SI")

# Plot for dist.gi
plot2 <- si_earlyR

library(cowplot)

plot_grid(plot1, plot2, ncol = 1)
```

#### Overlaying

```{r}
#turn epiestim SI into dataframe

si_data <- data.frame(si_distr = res_parametric_si$si_distr[0:11],
                      time = 0:10)
#plot 

plot <- ggplot() + 
  geom_line(data = si_df, aes(x = x, y = y, color = "earlyR"), size = 1) +
  geom_line(data = si_data, aes(x = time, y = si_distr, color = "EpiEstim"), size = 1) + 
  scale_color_manual(name = "Legend", values = c("EpiEstim" = "blue4", "earlyR" = "lightblue1")) +
  theme_bw() +
  labs(title = "Serial Interval",
       x = "Days after onset",
       y = "Probability")

plot
```

### R Estimation Analysis

```{r}
plot3 <- plot(res_parametric_si, "R")

# Plot for dist.gi
plot4 <- plot(res)

library(cowplot)

plot_grid(plot3, plot4, ncol = 1)
```

Given R is equal to 1.01 as shown on the second graph, it means that a person who already contracted the disease is likely to spread the disease to, at least, one other individual and the repective maximum likehood for that is about 4.5%.
