---
title: "EpiEstim x projections testing"
author: "Christine Sangphet"
date: "2024-07-13"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EpiEstim x projections

```{r}
#load environment 

library(EpiEstim)
library(ggplot2)
library(incidence)
library(outbreaks)
```

```{r}
#load data

data(Flu2009)
```

### SI estimation with EpiEstim

```{r}
res_parametric_si <- estimate_R(Flu2009$incidence, 
                                method="parametric_si",
                                config = make_config(list(
                                  mean_si = 2.6, 
                                  std_si = 1.5))
)
```

```{r}
plot(res_parametric_si, "SI")
```

### SI estimation with projections

```{r}
i <- (as.incidence(Flu2009$incidence$I, dates = Flu2009$incidence$dates))

plot(i) +
  theme_bw()
```

```{r}
#create a serial interval

library(distcrete)
library(epitrix)

mu <- 2.6
sigma <- 1.5
cv <- sigma / mu
params <- gamma_mucv2shapescale(mu, cv)
params
```

```{r}
si <- distcrete("gamma", shape = params$shape,
                scale = params$scale,
                interval = 1, w = 0.5)
si
```

```{r}
si_df <- data.frame(t = 0:10,
                    p = si$d(0:10))
ggplot(si_df, aes(x = t, y = p)) +
   theme_bw() +
  geom_col() +
  labs(title = "Serial interval",
       x = "Days after onset",
       y = "Probability")
```

### Side-by-side comparison of SI estimation

```{r}
library(cowplot)

plot1 <- plot(res_parametric_si,"SI")

plot2 <- ggplot(si_df, aes(x = t, y = p)) +
   theme_bw() +
  geom_col() +
  labs(title = "Serial interval",
       x = "Days after onset",
       y = "Probability")

plot_grid(plot1, plot2, ncol = 2)
```

### Overlaying comparison of SI estimation

```{r}
#turn epiestim SI into dataframe

si_data <- data.frame(si_distr = res_parametric_si$si_distr[0:11],
                      time = 0:10)

# Create the plot

plot <- ggplot(si_df, aes(x = t, y = p)) +
  geom_col(aes(fill = "projections")) + 
  geom_line(data = si_data, aes(x = time, y = si_distr, color = "EpiEstim"), size = 1) + 
  scale_fill_manual(name = "Legend", values = "lightblue2", labels = "Projections") +
  scale_color_manual(name = "Legend", values = "blue4", labels = "EpiEstim") +
  theme_bw() +
  labs(title = "Serial Interval",
       x = "Days after onset",
       y = "Probability")

plot
```

NOTE: projections uses other packages when estimating for SI (discrete, epitrix). projections specializes in projecting future incidences, which we are not testing in this part of the research. with that, the estimations of SI for epitrix and projections will be the same
