---
title: "epigrowthfit Testing"
author: "Christine Sangphet"
date: "2024-07-03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load environment

library(epigrowthfit)
```

# finalsize(): compute the expected epidemic

```{r}
R0 <- 10^seq(-3, 1, length.out = 151L)
plot(R0, finalsize(R0, S0 = 1, I0 = 0), type = "l", las = 1,
xlab = "basic reproduction number",
ylab = "final size")
```

# gi(): generation interval distribution

```{r}
latent <- c(0.026, 0.104, 0.182, 0.246, 0.318, 0.104,
0.013, 0.004, 0.003)
m <- length(latent)

infectious <- c(0.138, 0.462, 0.256, 0.078, 0.041, 0.007,
0.004, 0.004, 0.006, 0.004)
n <- length(infectious)
```

```{r}
#histogram of samples

y <- rgi(1e06, latent, infectious)
hist(y, breaks = seq(0, m + n + 1), freq = FALSE, las = 1,
ylab = "relative frequency",
main = "")
```

```{r}
#density and distribution functions

x <- seq(0, m + n + 1, by = 0.02)
fx <- dgi(x, latent, infectious)
Fx <- pgi(x, latent, infectious)
plot(x, fx, type = "l", las = 1, # consistent with histogram
xlab = "generation interval",
ylab = "density function")

plot(x, Fx, type = "l", las = 1,
xlab = "generation interval",
ylab = "distribution function")
```

```{r}
#quantile function
p <- seq(0, 1, by = 0.001)
qp <- qgi(p, latent, infectious)

plot(p, qp, type = "l", las = 1,
xlab = "probability",
ylab = "quantile function")
```

# plot.egf: plot nonlinear mixed effects models of epidemic growth

```{r}
example("egf", package = "epigrowthfit")
l <- list(legend = list(cex = 0.8),
value = list(cex = 0.8, font = 2),
ci = list(cex = 0.8))
control <- egf_control_plot(doubling = l)
op <- par(mar = c(3.5, 5, 5, 1))
plot(m1,
type = "interval",
show_predict = 2L,
show_doubling = 2L,
control = control)
```

```{r}
plot(m1,
type = "cumulative",
main = "Fitted exponential model",
sub = quote(paste("Country", country)))
par(op)
op <- par(mar = c(3.5, 9.5, 5, 1))
plot(m1, type = "rt", subset = quote(country %in% LETTERS[4:6]))
par(op)
```

# R0(): compute the basic reproduction number

```{r}
r <- seq(0, 1, 0.02)
breaks <- seq(0, 20, 1)
probs <- diff(pgamma(breaks, shape = 1, scale = 2.5))
plot(r, R0(r, breaks, probs), las = 1,
xlab = "initial exponential growth rate",
ylab = "basic reproduction number")
```

# simulate.egf(): simulation and parametric bootstrapping

### Simulates incidence data conditional on a fitted nonlinear mixed effects model of epidemic growth

```{r}
example("egf", package = "epigrowthfit")
zz <- simulate(m2, nsim = 6L, seed = 181952L, bootstrap = TRUE)
str(zz)
matplot(t(zz[["bootstrap"]][!m2[["random"]], ]),
type = "o", las = 1, xlab = "simulation", ylab = "value")
```

# simulate.egf_model(): simulating incidence time series

### Simulates equally spaced incidence time series according to a specified nonlinear model

```{r}
r <- 0.04
c0 <- 400
s <- 0.2
mu <- log(c(r, c0))
Sigma <- diag(rep.int(s^2, length(mu)))
zz <- simulate(object = egf_model(curve = "exponential", family = "pois"),
nsim = 20L,
seed = 202737L,
mu = mu,
Sigma = Sigma,
cstart = 10)

str(zz)

mm <- egf(zz)
(pp <- cbind(actual = coef(zz), fitted = coef(mm)))
```
