---
title: "EpiNow2 Testing"
author: "Christine Sangphet"
date: "2024-06-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reporting delays, incubation period, and generation time

```{r}
#load environment

library(EpiNow2)
```

```{r}
#if data on the delay between onset and infection was available we could fit a distribution to it, using estimate_delay(), with appropriate uncertainty as follows

reporting_delay <- estimate_delay(
  rlnorm(1000, log(2), 1),
  max_value = 14, bootstraps = 1
)
```

```{r}
#If data was not available we could instead specify an informed estimate of the likely delay using the distribution functions Gamma or LogNormal. To demonstrate, we choose a lognormal distribution with mean 2, standard deviation 1 and a maximum of 10

reporting_delay <- LogNormal(mean = 2, sd = 1, max = 10)
reporting_delay
```

```{r}
example_generation_time
```

```{r}
example_incubation_period
```

# function epinow()

### This function represents the core functionality of the package and includes results reporting, plotting, and optional saving. It requires a data frame of cases by date of report and the distributions defined above.

```{r}
#Load example case data

reported_cases <- example_confirmed[1:60]
head(reported_cases)
```

### Estimate cases by date of infection, the time-varying reproduction number, the rate of growth, and forecast these estimates into the future by 7 days.

### Summarize the posterior and return a summary table and plots for reporting purposes. Here we use the default model parameterisation that prioritizes real-time performance over run-time or other considerations.

```{r}
estimates <- epinow(
  data = reported_cases,
  generation_time = generation_time_opts(example_generation_time),
  delays = delay_opts(example_incubation_period + reporting_delay),
  rt = rt_opts(prior = list(mean = 2, sd = 0.2)),
  stan = stan_opts(cores = 4, control = list(adapt_delta = 0.99)),
  verbose = interactive()
)
names(estimates)
```

```{r}
#both summary measures and posterior samples are returned for all parameters in an easily explored format which can be accessed using summary

knitr::kable(summary(estimates))
```

```{r}
#summarised parameter estimates can also easily be returned, either filtered for a single parameter or for all parameters

head(summary(estimates, type = "parameters", params = "R"))
```

```{r}
#reported cases are returned in a separate data frame in order to streamline the reporting of forecasts and for model evaluation

head(summary(estimates, output = "estimated_reported_cases"))
```

```{r}
# a range of plots are returned (with the single summary plot shown below). These plots can also be generated using the following plot method

plot(estimates)
```

# Alternative way: single region

```{r}
#Here we use the example delay and generation time distributions that come with the package. This should be replaced with parameters relevant to the system that is being studied

options(mc.cores = 4)
reported_cases <- example_confirmed[1:60]
reporting_delay <- LogNormal(mean = 2, sd = 1, max = 10)
delay <- example_incubation_period + reporting_delay
rt_prior <- list(mean = 2, sd = 0.1)
```

```{r}
#We can then run the epinow() function with the same arguments as estimate_infections()

res <- epinow(reported_cases,
  generation_time = generation_time_opts(example_generation_time),
  delays = delay_opts(delay),
  rt = rt_opts(prior = rt_prior)
)
```

```{r}
res$plots$R
```

# **Running the model simultaneously on multiple regions**

```{r}
#say, for example, we construct a dataset containing two regions, testland and realland (in this simple example both containing the same case data)

cases <- example_confirmed[1:60]
cases <- data.table::rbindlist(list(
  data.table::copy(cases)[, region := "testland"],
  cases[, region := "realland"]
 ))
```

```{r}
#To then run this on multiple regions using the default options above, we could use

region_rt <- regional_epinow(
  data = cases,
  generation_time = generation_time_opts(example_generation_time),
  delays = delay_opts(delay),
  rt = rt_opts(prior = rt_prior)
)
```

```{r}
region_rt$summary$summarised_results$table
```

```{r}
region_rt$summary$plots$R
```

# function regional_epinow()

### The regional_epinow() function runs the epinow() function across multiple regions in an efficient manner.

```{r}
#Define cases in multiple regions delineated by the region variable

reported_cases <- data.table::rbindlist(list(
  data.table::copy(reported_cases)[, region := "testland"],
  reported_cases[, region := "realland"]
))
head(reported_cases)
```

```{r}
#Calling regional_epinow() runs the epinow() on each region in turn (or in parallel depending on the settings used). Here we switch to using a weekly random walk rather than the full Gaussian process model giving us piecewise constant estimates by week

estimates <- regional_epinow(
  data = reported_cases,
  generation_time = generation_time_opts(example_generation_time),
  delays = delay_opts(example_incubation_period + reporting_delay),
  rt = rt_opts(prior = list(mean = 2, sd = 0.2), rw = 7),
  gp = NULL,
  stan = stan_opts(cores = 4, warmup = 250, samples = 1000)
)
```

```{r}
#summary measures that are returned include a table formatted for reporting (along with raw results for further processing). Futures updated will extend the S3 methods used above to smooth access to this output

knitr::kable(estimates$summary$summarised_results$table)
```

```{r}
#a range of plots are again returned (with the single summary plot shown below)

estimates$summary$summary_plot
```

# function estimate_infections()

```{r}
#load environment 

library("rstan")
```

```{r}
#set the number of cores to use to 4 but the optimal value here will depend on the computing resources available

options(mc.cores = 4)
```

```{r}
#we will use an example data set that is included in the package, representing an outbreak of COVID-19 with an initial rapid increase followed by decreasing incidence

library("ggplot2")
reported_cases <- example_confirmed[1:60]
ggplot(reported_cases, aes(x =  date, y = confirm)) +
  geom_col() +
  theme_minimal() +
  xlab("Date") +
  ylab("Cases")
```

# 
