---
title: "epitrix Testing"
author: "Christine Sangphet"
date: "2024-06-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fitting a gamma distribution to delay data

### Simulate data which replicate the serial interval (SI), i.e. the delays between primary and secondary symptom onsets, in Ebola Virus Disease (EVD)

```{r}
#load environment

library(epitrix)
```

```{r}
#start by converting previously estimates of the mean and standard deviation of the SI to the parameters of a Gamma distribution 

mu <- 15.3 # mean in days days
sigma <- 9.3 # standard deviation in days
cv <- sigma/mu # coefficient of variation
cv
```

```{r}
param <- gamma_mucv2shapescale(mu, cv) # convertion to Gamma parameters
param

#the shape and scale are parameters of a Gamma distribution we can use to generate delays. However, delays are typically reported per days, which implies a discretization (from continuous time to discrete numbers).
```

```{r}
#we use the package distcrete to achieve this discretization. It generates a list of functions, including one to simulate data ($r), which we use to simulate 500 delays

library(distcrete)

si <- distcrete::distcrete("gamma", interval = 1,
               shape = param$shape,
               scale = param$scale, w = 0)
si
```

```{r}
set.seed(1)
x <- si$r(500)
head(x, 10)
```

```{r}
hist(x, col = "grey", border = "white",
     xlab = "Days between primary and secondary onset",
     main = "Simulated serial intervals")

#x contains simulated data, for illustrative purpose. In practice, one would use real data from an ongoing outbreaks
```

```{r}
#now we use fit_disc_gamma to estimate the parameters of a dicretised Gamma distribution from the data

si_fit <- fit_disc_gamma(x)
si_fit
```

# Converting a growth rate (r) to a reproduction number (R0)

### The package incidence can fit a log-linear model to incidence curves (function fit), which produces a growth rate (r). This growth rate can in turn be translated into a basic reproduction number (R0) using r2R0. We illustrate this using simulated Ebola data from the outbreaks package, and using the serial interval from the previous example.

```{r}
#load environment

library(incidence)
library(outbreaks)
```

```{r}
i <- incidence(ebola_sim$linelist$date_of_onset)
i
```

```{r}
f <- fit(i[1:150]) # fit on first 150 days
```

```{r}
plot(i[1:200], fit = f, color = "#9fc2fc")
```

```{r}
r2R0(f$info$r, si$d(1:100))
```

```{r}
r2R0(f$info$r.conf, si$d(1:100))
```

```{r}
#we can also use the function lm2R0_sample to generate samples of R0 values compatible with a model fit

R0_val <- lm2R0_sample(f$model, si$d(1:100), n = 100)
head(R0_val)
```

```{r}
hist(R0_val, col = "grey", border = "white")
```

# Estimate incubation periods

### The function empirical_incubation_dist() computes the discrete probability distribution by giving equal weight to each patient. Thus, in the case of N patients, the n possible exposure dates of a given patient get the overall weight 1/(n\*N). The function returns a data frame with column incubation_period containing the different incubation periods with a time step of one day and their relative_frequency

```{r}
#load environment 

library(magrittr)
library(epitrix)
library(distcrete)
library(ggplot2)
```

```{r}
#make a linelist object containing toy data with several possible exposure dates for each case

ll <- sim_linelist(15)

x <- 0:15
y <- distcrete("gamma", 1, shape = 12, rate = 3, w = 0)$d(x)
mkexposures <- function(i) {
  i - sample(x, size = sample.int(5, size = 1), replace = FALSE, prob = y)
}
exposures <- sapply(ll$date_of_onset, mkexposures)
ll$dates_exposure <- exposures

print(ll)
```

```{r}
#empirical distribution

incubation_period_dist <- empirical_incubation_dist(ll, date_of_onset, dates_exposure)
print(incubation_period_dist)
```

```{r}
ggplot(incubation_period_dist, aes(incubation_period, relative_frequency)) +
  geom_col()
```

```{r}
#fit discrete gamma

fit <- fit_gamma_incubation_dist(ll, date_of_onset, dates_exposure)
print(fit)
```

```{r}
x = c(0:10)
y = fit$distribution$d(x)
ggplot(data.frame(x = x, y = y), aes(x, y)) +
  geom_col(data = incubation_period_dist, aes(incubation_period, relative_frequency)) +
  geom_point(stat="identity", col = "red", size = 3) +
  geom_line(stat="identity", col = "red")
```
