---
title: "R0 Testing"
author: "Christine Sangphet"
date: "2024-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# est.GT: Find the best-fitting GT distribution for a series of serial interval

```{r}
#load environment 

library(R0)
```

```{r}
# Data taken from traced cases of H1N1 viruses.

data(H1N1.serial.interval)
est.GT(serial.interval=H1N1.serial.interval)
```

```{r}
#the same result can be achieved with two vectors of dates of onset.

# Here we use the same data, but trick the function into thinking onset dates are all "0".

est.GT(infector.onset.dates=rep(0,length(H1N1.serial.interval)),
infectee.onset.dates=H1N1.serial.interval)
```

# est.R0.AR: Estimate R0 from attack rate of an epidemic

### Woodall reported an attack rate of 0.31 in a population of 1732 during the 1957 H2N2 influenza pandemic (’Age and Asian Influenza, 1957’, BMJ, 1958)

```{r}
est.R0.AR(pop.size=1732, AR=0.31)
```

```{r}
est.R0.AR(AR=0.31)
```

```{r}
est.R0.AR(pop.size=1732, incid=31)
```

```{r}
est.R0.AR(pop.size=1732, incid=c(2,3,4,7,4,2,4,5))
```

```{r}
est.R0.AR(pop.size=1732, incid=c(2,3,0,7,4,2,0,5))
```

# est.R0.EG: Estimate R from exponential growth rate

### Data is taken from the paper by Nishiura for key transmission parameters of an institutional outbreak during 1918 influenza pandemic in Germany)

```{r}
data(Germany.1918)
mGT<-generation.time("gamma", c(3, 1.5))
est.R0.EG(Germany.1918, mGT, begin=1, end=27)
```

# est.R0.ML: Estimate the reproduction number by maximum likelihood

### Data is taken from paper by Nishiura for key transmission parameters of an institutional outbreak during the 1918 influenza pandemic in Germany)

```{r}
data(Germany.1918)
mGT<-generation.time("gamma", c(2.45, 1.38))
est.R0.ML(Germany.1918, mGT, begin=1, end=27, range=c(0.01,50))
```

```{r}
res=est.R0.ML(Germany.1918, mGT, begin=1, end=27, range=c(0.01,50))
plot(res)
```

# est.R0.TD: Estimate the time dependent reproduction number

### Data is taken from the paper by Nishiura for key transmission parameters of an institutional outbreak during 1918 influenza pandemic in Germany)

```{r}
data(Germany.1918)
mGT<-generation.time("gamma", c(3, 1.5))
TD <- est.R0.TD(Germany.1918, mGT, begin=1, end=126, nsim=100)
```

```{r}
TD
```

# estimate.R: Estimate R0 for one incidence dataset using several methods

```{r}
#Outbreak during 1918 influenza pandemic in Germany

data(Germany.1918)
mGT<-generation.time("gamma", c(3, 1.5))
estR0<-estimate.R(Germany.1918, mGT, begin=1, end=27, methods=c("EG", "ML", "TD", "AR", "SB"),
pop.size=100000, nsim=100)
attributes(estR0)
```

```{r}
#Estimates results are stored in the $estimates object

estR0
```

# generation.time: Generation Time distribution

```{r}
#GT for children at house(from Cauchemez PNAS 2011)

GT.chld.hsld1<-generation.time("empirical", c(0,0.25,0.2,0.15,0.1,0.09,0.05,0.01))
plot(GT.chld.hsld1, col="green")
GT.chld.hsld1
```

```{r}
GT.chld.hsld2<-generation.time("gamma", c(2.45, 1.38))
GT.chld.hsld2
```

```{r}
# T for school & community

GTs1 <- generation.time("empirical", c(0, 0.95, 0.05))
plot(GTs1, col='blue')
```

```{r}
plot(GT.chld.hsld1, ylim=c(0,0.5), col="red")
par(new=TRUE)
plot(GT.chld.hsld2, xlim=c(0,7), ylim=c(0,0.5), col="black")
```

# impute.incid: Optimization routine for incidence imputation

### Data is taken from the paper by Nishiura for key transmission parameters of an institutional outbreak during 1918 influenza pandemic in Germany

```{r}
#load environment

library(linelist)
library(epitrix)
```

```{r}
data(Germany.1918)
mGT<-generation.time("gamma", c(2.6,1))
```

```{r}
sen = sensitivity.analysis(sa.type="time", incid=Germany.1918, GT=mGT, begin=1:15, end=16:30,
est.method="EG")
```

```{r}
plot(sen, what=c("criterion","heatmap"))
```

# sim.epid: Epidemic outbreak simulation

### In this example we simulate n=100 epidemic curves, with peak value at 150 incident cases,

## and maximum epidemic length of 30 time units. Only the outbreak phase is computed. When the peak value is reached, the process is stopped and another epidemic is generated.

```{r}
sim.epid(epid.nb=100, GT=generation.time("gamma",c(3,1.5)), R0=1.5,
epid.length=30, family="poisson", peak.value=150)

#Here, a 30*100 matrix is returned. Each column is a single epidemic.
```
